A state space model for characterizing the trajectory dynamics of replay

### 2020-05-20 Article.tex
Update on Overleaf.

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v3}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} 19 simulated place cells that exhibit three different sequence patterns. For the first third of time, one cell fires repeatedly, representing one stationary location. For the second third of time, the cells fire in sequence, representing a smooth continuous trajectory through space. For the last third cells fire in an incoherent pattern, representing a non-smooth fragmented trajectory through space. \textbf{(B)} Like the Bayesian Decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} This is combined with an explicit model of each dynamic state--which determines how latent position can change based on the posterior from the previous time step. In the hover state, this is modeled as the identity matrix because the position remains the same. In the continuous state, the dynamics are modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close. In the fragmented state, the dynamics are modeled as uniform, meaning the next position is equally likely to be anywhere. Note that the dynamics displayed only reflect the within-state dynamics. \textbf{(D)} The probability of remaining in a particular state versus switching to another state. This is modeled as having a high probability of remaining in a particular state with a small probability of switching to one of the other dynamics at each time step. \textbf{(F)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time (lines). This can also be used to get an estimate of latent position over time (right panel). \textbf{(G)} The probability of each state given a particular constant speed trajectory. Each speed corresponds to a simulated spiking sequence at that speed. The lines correspond to the average probability of that state over the sequence. Dotted line represents the 0.8 classification threshold we use to classify each range of speeds and shaded regions correspond to the classification we have given the range of speeds.
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
In order to introduce and validate the model, we demonstrate how the model works on a simulated dataset (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then apply our decoding algorithm to the spiking sequence in Figure \ref{1}A. For the first third of this sequence, a single place cell fires repeatedly. For the middle third of the sequence, the cells fire in sequential spatial order. For the last third, the cells fire in a incoherent spatial order. These firing patterns represent three different types of sequence dynamics, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize replay events in terms of a mixture of these three dynamics at every time point.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v3}
\caption{Demonstration of the state space model on real data using both sorted and clusterless spikes on the same SWR. \textbf{(A)} Decoding using sorted spikes (arranged top to bottom). In the top panel, 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure 1F. Shaded regions correspond to the speed classifications as in Figure 1G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center wells respectively. The animal's position is indicated by the red triangle. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as(A).  \textbf{(C)} 1D MAP estimate of the latent position in (A) projected back into 2D. Color indicates time. The animal's position is denoted by the red circle. Light grey lines show the animal's 2D position over the entire epoch. L, R, and C correspond to the left, right and center well as in (A). \textbf{(D)} 1D MAP estimate of the latent position in (B) projected back into 2D. Figure conventions are the same as in (C).
}
\label{2}
\end{figure*}
\subsection*{The model finds known continuous replays with sorted and clusterless spikes}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v3}
\caption{A-D. Four examples of replays that would not be well-characterized by using the Bayesian decoder. Figure conventions are the same as in Figure 2. \textbf{(A)} A replay that starts down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A replay that stays at the left well for the entirety of the SWR while the animal is in the center arm. \textbf{(C)} A replay that begins stationary at the left well and then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A replay that begins stationary at the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} Percentage of SWRs in each epoch that has some content that falls into one of the categories (Hover, Hover-Continuous-Mix, Continuous, Fragmented-Continuous-Mix, or Fragmented). \textbf{(F)} Of the classified content, the percentage of SWRs in each epoch that have some Hover, Hover-Continuous-Mix, or Continuous content. \textbf{(G)} Of the classified content, the percentage of SWR in each epoch that has some Continuous content.
}
\label{3}
\end{figure*}

@@

\subsection*{Many replay trajectories are closer to the animal's movement speed}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v3}
\caption{
Analysis of SWRs from all animals. \textbf{(A)} UpSet plot---which is similar to a Venn diagram with more than 3 sets---of the most common sets of dynamics within each ripple. Each row represents a possible category present in the SWR. Each column represents a set of dynamics categories, where filled-in black dots with an edge between the dots indicates that multiple categories are present in the SWR (at temporally distinct times). The sets are ordered by how often they occurred as indicated by the bar plot above each category. The total number of each category is indicated by the rotated bar plot on the left. \textbf{(B)} Distribution of the duration of each dynamic category within a SWR. \textbf{(C)} The distance of the average latent position from the animal's position for each dynamic category within each SWR. \textbf{(D)} Position of hovers on the W-track at least 30 cm away from the animal's position. \textbf{(E)} Distribution of multiunit spike rates for each dynamic category within each SWR.
}
\label{4}
\end{figure*}

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v1}
\caption{Comparison of real vs. position shuffled data for two epochs from different animals. Line represents the percent of SWR classified in that epoch for real data. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that epoch, destroying position information but preserving spiking timing.}
\label{Figure5}
\end{figure}
\section*{Discussion}

@@

\subsection*{Classification of Dynamics}
We used a threshold of 0.8 to classify the probability of each state into 5 categories. Time periods during sharp wave ripples are labeled as Hover, Continuous, or Fragmented when the probability of each state is above 0.8. Time periods are labeled as Hover-Continuous-Mix or Fragmented-Continuous-Mix when the sum of Hover and Continuous or Fragmented and Continuous are above 0.8, respectively. Time periods where none of these criterion are met are considered unclassified because there is not an interpretation for all three dynamics being equally likely.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Supplementary Information %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\beginsupplement
\captionsetup*{format=largeformat}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1-supplemental1/Figure1_v1_supplemental1}
\caption{The model is robust to changes in the discrete transition matrix. Each plot shows the probability of each dynamic on simulated data example from Figure 1 with a different diagonal value--which governs the probability of remaining in that state. The off-diagonal values--the probability of switching to one of the other dynamics--are set to be equally likely with the remainder of the probability, as in Figure 1D. The first plot (upper left) shows the case when all dynamics are equally likely. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero--i.e. the case where there is no probability of switching to another state.}
\label{fig:Figure1-Figure supplement 1}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2-supplemental1/Figure2_v2-supplemental1}
\caption{Decoding the same SWR in Figure 2 using 2D clusterless and sorted spikes. \textbf{(A)} The left plot shows the spikes from cells arranged by the linear position of the peak of place field as in Figure 2. The middle plot shows the probability of each dynamic over time from the 2D decode. Shaded regions indicate classification category as in Figure 1G and 2. The rightmost plot shows the MAP estimate of the latent position with color indicating time. The latent position posterior summed over time is shown in purple underneath. The light grey lines represent the position of the animal over the entire epoch and the red dot represents the animal's position. \textbf{(B)} Same as in A, but with clusterless decoding.}
\label{fig:Figure2-Figure supplement 1}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2-supplemental2/Figure2_v2-supplemental2}
\caption{A-C. Examples of continuous replays. Left panel uses the same conventions as Figure 2A and 2B. Right panel shows the 1D MAP estimate projected back to 2D as in Figure 2C. Color indicates time. Light grey lines indicate the animal's position over the entire epoch.}
\label{fig:Figure2-Figure supplement 2}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3-supplemental1/Figure3_v2_supplemental1}
\caption{More examples of replays that would not be well-characterized by using the Bayesian decoder. Conventions are the same as in Figure 3. \textbf{(A)} A continuous replay that starts in the left arm back toward the center well, jumps to the right arm and continuous back toward the center well. \textbf{(B)} A continuous replay that travels down the right arm toward the center well, proceeds past the choice point toward the left well and then returns back to the choice point. \textbf{(C)} A replay that starts down the left arm toward the center arm and turns into a hover near the choice point. \textbf{(D)} The first half is a fragmented replay and the second half is unclassified. \textbf{(E)} A hover in the first half of the SWR on the right arm and then a hover in the SWR at the left well. \textbf{(F)} A hover that starts at the animal's position at the center well, then jumps to the left well and comes back toward the center well.
}
\label{fig:Figure3-Figure supplement 1}
\end{figure*}

>>>>

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v4}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} 19 simulated place cells that exhibit three different sequence patterns. For the first third of time, one cell fires repeatedly, representing one stationary location. For the second third of time, the cells fire in sequence, representing a smooth continuous trajectory through space. For the last third cells fire in an incoherent pattern, representing a non-smooth fragmented trajectory through space. \textbf{(B)} Like the Bayesian Decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} This is combined with an explicit model of each dynamic state--which determines how latent position can change based on the posterior from the previous time step. In the stationary state, this is modeled as the identity matrix because the position remains the same. In the continuous state, the dynamics are modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close. In the fragmented state, the dynamics are modeled as uniform, meaning the next position is equally likely to be anywhere. Note that the dynamics displayed only reflect the within-state dynamics. \textbf{(D)} The probability of remaining in a particular state versus switching to another state. This is modeled as having a high probability of remaining in a particular state with a small probability of switching to one of the other dynamics at each time step. \textbf{(F)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time (lines). This can also be used to get an estimate of latent position over time (right panel). \textbf{(G)} The probability of each state given a particular constant speed trajectory. Each speed corresponds to a simulated spiking sequence at that speed. The lines correspond to the average probability of that state over the sequence. Dotted line represents the 0.8 classification threshold we use to classify each range of speeds and shaded regions correspond to the classification we have given the range of speeds.
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
In order to introduce and validate the model, we demonstrate how the model works on a simulated dataset (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then apply our decoding algorithm to the spiking sequence in Figure \ref{1}A. For the first third of this sequence, a single place cell fires repeatedly. For the middle third of the sequence, the cells fire in sequential spatial order. For the last third, the cells fire in a incoherent spatial order. These firing patterns represent three different types of sequence dynamics, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize replay events in terms of a mixture of these three dynamics at every time point.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v4}
\caption{Demonstration of the state space model on real data using both sorted and clusterless spikes on the same SWR. \textbf{(A)} Decoding using sorted spikes (arranged top to bottom). In the top panel, 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure 1F. Shaded regions correspond to the speed classifications as in Figure 1G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center wells respectively. The animal's position is indicated by the red triangle. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as(A).  \textbf{(C)} 1D MAP estimate of the latent position in (A) projected back into 2D. Color indicates time. The animal's position is denoted by the red circle. Light grey lines show the animal's 2D position over the entire epoch. L, R, and C correspond to the left, right and center well as in (A). \textbf{(D)} 1D MAP estimate of the latent position in (B) projected back into 2D. Figure conventions are the same as in (C).
}
\label{2}
\end{figure*}
\subsection*{The model finds known continuous replays with sorted and clusterless spikes}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v4}
\caption{A-D. Four examples of replays that would not be well-characterized by using the Bayesian decoder. Figure conventions are the same as in Figure 2. \textbf{(A)} A replay that starts down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A replay that stays at the left well for the entirety of the SWR while the animal is in the center arm. \textbf{(C)} A replay that begins stationary at the left well and then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A replay that begins stationary at the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} Percentage of SWRs in each epoch that has some content that falls into one of the categories (Stationary, Stationary-Continuous-Mix, Continuous, Fragmented-Continuous-Mix, or Fragmented). \textbf{(F)} Of the classified content, the percentage of SWRs in each epoch that have some Stationary, Stationary-Continuous-Mix, or Continuous content. \textbf{(G)} Of the classified content, the percentage of SWR in each epoch that has some Continuous content.
}
\label{3}
\end{figure*}

@@

\subsection*{Many replay trajectories are closer to the animal's movement speed}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v4}
\caption{
Analysis of SWRs from all animals. \textbf{(A)} UpSet plot---which is similar to a Venn diagram with more than 3 sets---of the most common sets of dynamics within each ripple. Each row represents a possible category present in the SWR. Each column represents a set of dynamics categories, where filled-in black dots with an edge between the dots indicates that multiple categories are present in the SWR (at temporally distinct times). The sets are ordered by how often they occurred as indicated by the bar plot above each category. The total number of each category is indicated by the rotated bar plot on the left. \textbf{(B)} Distribution of the duration of each dynamic category within a SWR. \textbf{(C)} The distance of the average latent position from the animal's position for each dynamic category within each SWR. \textbf{(D)} Position of stationary trajectories on the W-track at least 30 cm away from the animal's position. \textbf{(E)} Distribution of multiunit spike rates for each dynamic category within each SWR.
}
\label{4}
\end{figure*}

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v2}
\caption{}
\label{Figure5}
\end{figure*}
\section*{Discussion}

@@

\subsection*{Classification of Dynamics}
We used a threshold of 0.8 to classify the probability of each state into 5 categories. Time periods during sharp wave ripples are labeled as Stationary, Continuous, or Fragmented when the probability of each state is above 0.8. Time periods are labeled as Stationary-Continuous-Mix or Fragmented-Continuous-Mix when the sum of Stationary and Continuous or Fragmented and Continuous are above 0.8, respectively. Time periods where none of these criterion are met are considered unclassified because there is not an interpretation for all three dynamics being equally likely.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Supplementary Information %
%%%%%%%%%%%%%%%%%%%%x%%%%%%%%%
\beginsupplement
\captionsetup*{format=largeformat}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1-supplemental1/Figure1_v2_supplemental1}
\caption{The model is robust to changes in the discrete transition matrix. Each plot shows the probability of each dynamic on simulated data example from Figure 1 with a different diagonal value--which governs the probability of remaining in that state. The off-diagonal values--the probability of switching to one of the other dynamics--are set to be equally likely with the remainder of the probability, as in Figure 1D. The first plot (upper left) shows the case when all dynamics are equally likely. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero--i.e. the case where there is no probability of switching to another state.}
\label{fig:Figure1-Figure supplement 1}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2-supplemental1/Figure2_v3-supplemental1}
\caption{Decoding the same SWR in Figure 2 using 2D clusterless and sorted spikes. \textbf{(A)} The left plot shows the spikes from cells arranged by the linear position of the peak of place field as in Figure 2. The middle plot shows the probability of each dynamic over time from the 2D decode. Shaded regions indicate classification category as in Figure 1G and 2. The rightmost plot shows the MAP estimate of the latent position with color indicating time. The latent position posterior summed over time is shown in purple underneath. The light grey lines represent the position of the animal over the entire epoch and the red dot represents the animal's position. \textbf{(B)} Same as in A, but with clusterless decoding.}
\label{fig:Figure2-Figure supplement 1}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2-supplemental2/Figure2_v3-supplemental2}
\caption{A-C. Examples of continuous replays. Left panel uses the same conventions as Figure 2A and 2B. Right panel shows the 1D MAP estimate projected back to 2D as in Figure 2C. Color indicates time. Light grey lines indicate the animal's position over the entire epoch.}
\label{fig:Figure2-Figure supplement 2}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3-supplemental1/Figure3_v3_supplemental1}
\caption{More examples of replays that would not be well-characterized by using the Bayesian decoder. Conventions are the same as in Figure 3. \textbf{(A)} A continuous replay that starts in the left arm back toward the center well, jumps to the right arm and continuous back toward the center well. \textbf{(B)} A continuous replay that travels down the right arm toward the center well, proceeds past the choice point toward the left well and then returns back to the choice point. \textbf{(C)} A replay that starts down the left arm toward the center arm and turns into a stationary trajectory near the choice point. \textbf{(D)} The first half is a fragmented replay and the second half is unclassified. \textbf{(E)} A stationary trajectory in the first half of the SWR on the right arm and then a stationary trajectory in the SWR at the left well. \textbf{(F)} A stationary trajectory that starts at the animal's position at the center well, then jumps to the left well and comes back toward the center well.
}
\label{fig:Figure3-Figure supplement 1}
\end{figure*}

\begin{figure}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v1}
\caption{Comparison of real vs. position shuffled data for two epochs from different animals. Line represents the percent of SWR classified in that epoch for real data. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that epoch, destroying position information but preserving spiking timing.}
\label{Figure3-Figure supplement 2}
\end{figure}
